---
title: "Project Sample E"
description: |
  Wordcloud and sentiment analysis of one of my favorite books!
author:
  - name: Taylor
    url: {}
date: 2021-03-15
output:
  distill::distill_article:
    self_contained: false
    code_folding: true
---


```{r setup, include=FALSE, message=FALSE, warning=FALSE, results=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(tidytext)
library(textdata)
library(pdftools)
library(ggwordcloud)
library(jpeg)
library(grid)
library(ggimage)
library(RColorBrewer)
```

```{r, cache = TRUE}
# Read in the pdf
eragon_text <- pdf_text(here("data", "Eragon.pdf"))
```

## 1. Wordcloud of *Eragon* text

```{r, results=FALSE}
# make the data tidy
eragon_tidy <- data.frame(eragon_text) %>% 
  mutate(text_full = str_split(eragon_text, pattern = "\\n")) %>% 
  unnest(text_full)%>% 
  mutate(text_full = str_trim(text_full))

# Make it into a dataframe
eragon_df <- eragon_tidy %>% 
  slice(-(1:12))

# Create tokens
eragon_tokens <- eragon_df %>% 
  unnest_tokens(word, text_full) %>% 
  dplyr::select(-eragon_text)

# Add up the wordcounts
eragon_wordcount <- eragon_tokens %>% 
  count(word)

# Remove all stopwords
eragon_nonstopwords<- eragon_tokens %>% 
  anti_join(stop_words, by = "word")

# Make a subset with counts of only non-stopwords
non_stop_counts <- eragon_nonstopwords %>% 
  count(word) %>% 
  filter(!(str_detect(word, "\\d"))) # Use this line to remove all rows of numbers

# Make a subset for a wordcloud!
eragon_top100 <- non_stop_counts %>% 
  arrange(-n) %>% 
  filter(word != "eragon") %>% 
  filter(word != "brom") %>%
  filter(word != "murtagh") %>%
  filter(word != "saphira") %>%
  slice(1:100)

# Read in data for background:
egg_img <- readJPEG(here("data", "egg2.jpg"))

# Make a wordcloud!
eragon_cloud <- ggplot(data = eragon_top100, aes(label = word, color = n))+
  annotation_custom(rasterGrob(egg_img,
                               width = unit(1, "npc"),
                               height = unit(1, "npc")),
                    -Inf, Inf, -Inf, Inf)+
  geom_text_wordcloud(aes(size =  n))+
  scale_color_gradient(low = "green", high = "white")+
  scale_size_area(max_size = 10)

eragon_cloud
```



**Image 1:** Illustrated word cloud for the entire text of *Eragon* (excluding main character names). Apologies for the low quality background image, I refused to use a screenshot from the movie. 



## 2. Sentiment analysis of *Eragon*

```{r, results=FALSE}
# Perform sentiment analysis using nrc

eragon_nrc <- eragon_nonstopwords %>%
  inner_join(get_sentiments("nrc"), by = "word")

eragon_nrc_counts <- eragon_nrc %>% 
  count(sentiment)
  

ggplot(data = eragon_nrc_counts, aes(reorder(sentiment, -n),n), fill = sentiment)+
  geom_col(aes(fill = sentiment))+
  coord_flip()+
  theme_bw()+
  labs(x = "Sentiment",
       y = "Count")+
  scale_fill_brewer(palette = "Set3")+
  theme(legend.position = "none")
  
```

**Image 2:** Sentiment analysis of the entire *Eragon* text. We can see that "negative," "positive," and "fear" are the sentiments with the highest counts in the book, while "disgust" and "surprise" are both close in last place. 


**Source:** Paolini, C. (2003). Eragon. New York: Alfred A. Knopf. Accessed through Academia.edu
