---
title: "Project Sample A"
description: |
  An exploration of cetacean species richness on the California coast.
author:
  - name: Taylor
    url: {}
date: 2021-03-15
output:
  distill::distill_article:
    self_contained: false
    code_folding: true
---


```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(raster)
library(tidyverse)
library(here)
library(sf)
library(fasterize)
library(rnaturalearth)
```

```{r}
# Read in the data all together
species <- here("data","ca_cetaceans", "ca_cetaceans")
species_files <- dir(species, full.names = TRUE, pattern = "*.tif")

# Rasterize all these files together using raster::stack
cetaceans_data <- raster::stack(species_files)

# Write a function to determine if species are present in a cell, with a threshold of 0.6 meaning "present"
is_present <- function(x, thresh = .6){
  y <- ifelse(x >= thresh, 1, 0)
  return(y)
}

# Apply the threshold function to our stack
species_richness <- calc(cetaceans_data, fun = is_present)

# Find out how many species are in each cell
species_richness1 <- calc(species_richness, fun = sum, na.rm = TRUE)

```

```{r, message = FALSE, warning=FALSE, results='hide'}
# Now create the file to use for the CA coastline

states <- ne_download(scale = 110, type = "states", category = "cultural", returnclass = "sf")

ca_state <- states %>% 
  filter(name == "California")
```

```{r}
# Crop the species raster to match the CA coastline
species_raster_cropped <- crop(species_richness1, extent(ca_state))

# Now turn this cropped raster into a dataframe
species_richness_df <- raster::rasterToPoints(species_raster_cropped) %>%
  as.data.frame() %>% 
  filter(layer != 0) # Filter out cells that have no data in them

# Now we have the species richness dataframe with which to make our ggplot!
```


```{r, fig.dim = c(8,6)}
# Now plot everything together

ggplot()+
  geom_raster(data = species_richness_df, aes(x = x, y = y, fill = layer))+
  geom_sf(data = ca_state, fill = "lemonchiffon2")+
  scale_fill_gradient(low = "white", high = "mediumblue", name = "Likely number of cetacean species present")+
  theme_minimal()+
  theme(panel.background = element_rect(fill = "grey90"))+
  labs( x = "Longitude",
        y = "Latitude",
        title = "Cetacean species richness on the California coast")
```



**Figure 1:** Map of species richness of cetaceans off the coast of California. Species richness is defined as the number of species likely to be present in a certain area. For a species to be considered "present," a likelihood threshold of 0.6 was observed, meaning that there had to be a minimum 60% chance that a species would be present in order for it to be considered "present" in our analysis. 


*Note:* I purposely did not clip the extent of the map because I liked the aesthetics of seeing the whole state shape rather than just the extent covered by the species raster. To clip I would have inserted a line in my ggplot of `coord_sf(xlim = c(-125, -115), ylim = c(32, 38))`.

**Sources:** 

- Kaschner, K., Rius-Barile, J., Kesner-Reyes, K., Garilao, C., Kullander, S., Rees, T., & Froese, R. (2016). AquaMaps: Predicted range maps for aquatic species. www.aquamaps.org
- Made with Natural Earth. Free vector and raster map data @ naturalearthdata.com.

##### End Project Sample A