---
title: "Ocean Conditions Dashboard"
description: |
  At-a-glance view of conditions for paddling in Santa Barbara, CA
author:
  - name: Taylor
    url: {}
date: 2025-02-25
output:
  distill::distill_article:
    self_contained: false
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(httr)
library(jsonlite)
library(tidyverse)
library(ggplot2)
library(DT)
library(gt)
library(kableExtra)
```

# Current Wind Conditions

Courtesty of windy.com

```{r wind_data, results='asis', echo=FALSE}
windy_iframe <- '<iframe width="650" height="450" src="https://embed.windy.com/embed2.html?lat=34.42&lon=-119.70&zoom=10&level=surface&overlay=wind" frameborder="0"></iframe>'
cat(windy_iframe)
```

# Swell and Wave Height

Wave height based on NOAA/NDBC* Santa Barbara Buoy 46054

*NDBC = National Data Buoy Center


```{r swell_data, echo=FALSE, message = FALSE, warning=FALSE}
# Define column names based on the NDBC format
col_names <- c("YY", "MM", "DD", "hh", "mm", "WDIR", "WSPD", "GST", "WVHT", 
               "DPD", "APD", "MWD", "PRES", "ATMP", "WTMP", "DEWP", "VIS", "PTDY", "TIDE")

# Read the data while skipping the first row (commented headers) and treating "MM" as NA
buoy_url <- "https://www.ndbc.noaa.gov/data/realtime2/46054.txt"
buoy_data <- read_table(buoy_url, col_names = col_names, skip = 2, na = "MM")

# Create a datetime column
buoy_data <- buoy_data %>%
  mutate(datetime = as.POSIXct(paste(YY, MM, DD, hh, mm), format="%Y %m %d %H %M", tz = "UTC"))

# Select relevant columns for display
buoy_data_clean <- buoy_data %>%
  select(datetime, WVHT, DPD, MWD, WTMP)  # Wave height, swell period, swell direction, water temp

# Display cleaned dataset
# print(head(buoy_data_clean))
```

```{r buoy_metrics, echo=FALSE, message=FALSE, warning=FALSE}

# # Ensure proper NA handling
# buoy_data_clean <- buoy_data_clean %>%
#   mutate(across(where(is.numeric), ~ replace_na(., NA_real_))) 

# Select the latest available data
latest_buoy_data <- buoy_data_clean %>%
  arrange(desc(datetime)) %>%
  slice(1:3) %>%
  mutate(
    datetime = with_tz(datetime, tzone = "America/Los_Angeles"),
    datetime = format(datetime, "%Y-%m-%d %H:%M"))

# Rename columns for display
colnames(latest_buoy_data) <- c("Date & Time", "Wave Height (m)", "Dominant Period (s)", 
                                "Mean Wave Direction (degrees)", "Water Temp (degrees C)")

kable(latest_buoy_data, format = "html", table.attr = "class='table table-striped'") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))
```

## Wave height over time

```{r message=FALSE, warning=FALSE, echo=FALSE}
# Prepare data: Filter recent data and ensure datetime is in Pacific Time
buoy_data_clean <- buoy_data_clean %>%
  mutate(datetime = as.POSIXct(datetime, tz = "UTC")) %>%  # Convert to POSIXct
  mutate(datetime = format(datetime, tz = "America/Los_Angeles", usetz = TRUE))  # Convert to PT

wave_height <- buoy_data_clean %>% 
  filter(datetime >= Sys.time() - 7*24*60*60) %>% 
  mutate(datetime = as.POSIXct(datetime, tz = "America/Los_Angeles"))

# Plot wave height over time
ggplot(data=subset(wave_height, !is.na(WVHT)), aes(x = datetime, y = WVHT)) +
  geom_point(color = "#264653", size = 2) +  # Highlight data points
  theme_light() +
  labs(
    title = "Wave Height Over Time",
    x = "Time (Pacific Time)",
    y = "Wave Height (m)"
  ) +
  scale_x_datetime(
    date_breaks = "6 hours", # Show labels every 4 hours
    date_labels = "%b %d\n%H:%M"  # Format: "Feb 25\n12:00"
  )
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for clarity
  )
```



# Visibility
```{r}

```

# Tides
```{r}

```

# Water Quality

I am still trying to find a good source of real-time data for this section! All I can find currently are local agencies that publish pdf reports of their findings, which would require web scraping skills that I have not quite developed yet. Let me know if you know of a source!

```{r water_quality}
# Placeholder for EPA/local agency data
# Example: Fetch and display water quality reports
```


# Summary View

```{r summary_table}
# Placeholder for formatted summary (tables, icons, key metrics)
```


